---
title: "Variability"
author: "Rogelio Caballero"
date: "May 5, 2017"
output: html_document
---
When sampling a random variable one can take different values for the number of trials ```nosim``` (for "number of simulations") and the amount of values that one collects in each trial ```n```.

The following generates the distributions of the means of ```n``` uniformly distributed numbers while keeping the number of trials ```nosim``` constant.

```{r}
df <- data.frame(m = vector(), sd = vector(), n = vector())

nosim <- 1000

for(n in c(5, 10, 50, 100)){
        m <- matrix(sample(1:6, nosim*n, replace = TRUE), nosim)
                
        sd <- apply(m, 1, sd)
        m <- apply(m, 1, mean)
        n1 <- rep(n, times = length(m))
                
        df1 <- data.frame(m, sd, n1)
        df <- rbind(df, df1)
}
```

The number of values over which the mean is taken in each trial has an effect on the spread of the distribution. The bigger ```n``` gets, the sharper the distribution.

```{r}
library(ggplot2)
h <- ggplot(df, aes(df$m)) + geom_histogram(binwidth = 0.2) +
        facet_grid(n1 ~ ., labeller = label_both)

h
```

However, if we fix ```n```, we see that variations in the number of simulations trivially only affect the height of the bars:

```{r}
df2 <- data.frame(m = vector(), sd = vector(), nosim = vector())

n <- 10

for(nosim in c(50, 100, 150)){
        m <- matrix(sample(1:6, nosim*n, replace = TRUE), nosim)
        
        sd <- apply(m, 1, sd)
        m <- apply(m, 1, mean)
        nosim1 <- rep(nosim, times = length(m))
        
        df1 <- data.frame(m, sd, nosim1)
        df2 <- rbind(df2, df1)
}

h1 <- ggplot(df2, aes(df2$m)) + geom_histogram(binwidth = 0.2) +
        facet_grid(nosim1 ~ ., labeller = label_both)

h1
```